AttCRISPR : an spacetime interpretable model for sgRNA efficiency prediction

### 2020-10-06 Paper/main.tex
10-6 paper

\text{\it y}=Score(\it X)\label{eq:01}\vspace*{-10pt}
\end{equation}\\

@@

dy = \sum^{d}_ia_idx_i\label{eq:02}\vspace*{-10pt}

@@

a_i=\frac{\partial}{\partial x_i} Score(X)\label{eq:03}\vspace*{-10pt}
\end{equation}\\

@@

y=\sum_i^lW_iA_i\cdot X_i\label{eq:04}\vspace*{-10pt}
\end{equation}\\

@@

\tilde{A}=BA\label{eq:05}\vspace*{-10pt}
\end{equation}\\

@@

\tilde{A}_i=\sum_j^lB_{ij}A_j\label{eq:06}\vspace*{-10pt}
\end{equation}\\

@@

d\tilde{A}_i=\sum_j^lB_{ij}EdA_j\label{eq:07}\vspace*{-10pt}
\end{equation}\\
 where $E$ is the identity matrix, and we define $B$ as the second-order preference matrix, 
 it can explain how a particular pattern containing two nucleotides affects the base sequence, 
 for example, how does the degree of preference for the G adjacent to PAM be affected by a T at 2 bases upstream?
 The previous work (even based on conventional machine learning algorithms with high interpretability) 
 was limited to dimers which can only work on the adjacent base pairs \citep{Liu2019} 
 or have to perform complex feature engineering \citep{MuhammadRafid2020}.
 In light of the above, we believe it is essential to develope a model which can not only match deep learning based model in performance, 
 but also be comparable to conventional machine learning algorithms in interpretability.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[scale=0.45]{category.png}}
\caption{Two categories of the deep learning models used in sgRNA related task. 
(a) Model work in spatial domain. (b) Model work in temporal domain.}\label{fig:01}
\end{figure}

@@

Attention mechanism has demonstrated its power in Natural Language Processing, Statistical Learning, Speech and Computer Vision \citep{chaudhari2019an}. 
 It makes model tends to focus selectively on parts of the input which is help in performing the task effectively. 
 Previous observation have shown that Cas9 preferentially binds sgRNAs containing purines but not pyrimidines \citep{wang2014genetic} and multiple thymine in the spacer impairing sgRNA activity \citep{wu2014genome-wide}, 
 that is to say some specific nucleotide and base position need more attention compared to others. 
 The above is the premise of introducing attention mechanism. Strictly speaking, we are not the first to bring attention mechanisms into this field. 
 The most similar approach to ours is the work based on transformer by \citeauthor{Liu2019}. 
 They use transformer, a components based on attention mechanism, instead of RNN to improve the ability of temporal feature extraction, hence, enhance the performance of their model \citep{Liu2019,vaswani2017attention}. 
 In our work,the interpretability benefit from attention mechanism is more focused.

In this paper, our main contributions are as follows:\vspace*{1pt}
\begin{itemize}
    \item Present a novel deep-learning-based model, which can extract potential feature representation of sgRNA sequence in both spatial and temporal domain parallelly. 
    Finally, the ensemble learning method is used to combine the two to achieve better performance than other models.
    It doesn't belong to any of the above categories of existing approaches.
    \item Introduce attention mechanism into our model. 
    As a result, It does not need post hoc explanations techniques based on input perturbation to explain itself. 
    It is intrinsic interpretable in both temporal and spatial domains.
    In the spatial domain it's at the nucleotide level, in our word, first-order preference, 
    while at the overall level in the temporal domain, in our word, second-order preference. 
    Thus, it is transformed from a black box to an intrinsically interpretable model with the performance of deep learning based model. 
    \item Throught abation analysis and testing a series of possible network structure, 
    we find there are multiple components and strategy can improve the performance of our model and constructed AttCRISPR, 
    which could outperform the current state-of-the-art tool on DeepHF dataset.\vspace*{1pt}
\end{itemize}

@@

To facilitate the training process, we can mapping each one-hot vector into a dense real-valued high-dimensional space, which is called embedding. 
Concretely speaking, at the matrix level, the formula is as follows
\begin{equation}
X_e=X_{oh}E_m\label{eq:08}\vspace*{-10pt}
\end{equation}\\
where $X_e$ named embedding matrix, $E_m\in\mathbb{R}^{4\times m}$ is a trainable transformational matrix, $m$ refers to the dimension of embedding space. 
We believe it is also meaningful to regard nucleotides in the sgRNA sequence as word, and the sgRNA sequence itself as a sentence, 
Guided by this belief $E_m$ is the word embedding matrix and $X_e$ is the sentence embedding in natural language processing (NLP).
Therefore, $X_e$ is used as an input of recurrent neural network (or its variant), which performs well in the NLP field.
In the area of NLP, there have been many methods based on pre-training and unsupervised learning been developed to determine the $E_m$. 
However, we believe that sgRNA sequence is not a natural language in the true sense, the method above does not apply to it. 
So $E_m$ will be trained along with the model as a whole.

@@

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[scale=0.15]{CNNv2.png}}
    \caption{The architecture of spatial domain method in AttCRISPR. 
    The input of the method is encoded sgRNA sequence, a $21\times 4$ one-hot matrix. 
    Then refine it through a spatial attention module, which could tell us the importance of a specific matrix element (or just say, pixel).
    Finally, a simple convolutional neural network is applied.}\label{fig:02}
\end{figure}
As Figure~2\vphantom{\ref{fig:02}} show, method in spatial domain relies on the convolutional neural network. 
As previously mentioned, sgRNA sequence has been encoded into a $21\times 4$ one-hot matrix $X_{oh}$, and we regard $X_{oh}$ as a binary image. 
Then, convolution kernels with different size are used to extract potential spatial features just like what other do in computer vision. 
According to the foregoing, the Spatial Attention Module proposed by \citeauthor{woo2018cbam:} can be applied in our method in spatial domain, 
which was used to improve the performance of CNN in vision tasks.

\begin{figure}[!tpb]%figure3
    \centerline{\includegraphics[scale=0.37]{spatialmodule.png}}
    \caption{Details of spatial attention module. A convolution layer is used to generate multi-channel map from $X_{oh}$.
    Then concatenated the output of both max-pooling and average-pooling method and forward it to the last convolution layer.
    A sigmoid is used to map the final result to a range of zero to one.
    }\label{fig:03}
\end{figure}

As Figure~3\vphantom{\ref{fig:03}} show, Given an one-hot map $X_{oh}$, spatial attention module generating spatial attention matrix 
$M_s\in\mathbb{R}^{l\times 4}$ with the same as $X_{oh}$ in shape. 
Each element of $M_s$ is constrained to a range of zero to one, implemented by a sigmoid function, which reflects the importance of the corresponding elements of $X_{oh}$.
The overall spatial attention process can be summarized as:
\begin{equation}
X_{mc} = f^{3\times4}(X_{oh})\label{eq:09}
\end{equation}
\begin{equation}
M_s = \sigma(f^{3\times2}([AvgPool(X_{mc});MaxPool(X_{mc})]))\label{eq:10}
\end{equation}
\begin{equation}
X_{rf} = M_s\otimes X_{oh}\label{eq:11}\vspace*{-10pt}
\end{equation}\\

where $f^{p\times q}$ represents a convolution operation with the filter size of $p\times q$, 
$p,q\in\mathbb{Z}^{+}$, $X_{mc}$ is a multi-channel map generated by $X_{oh}$, $\sigma$ denotes the sigmoid function, 
AvgPool denotes the average-pooling operation, MaxPool denotes the max-pooling operation, $\otimes$ denotes element-wise multiplication.

Spatial attention matrix $M_s$ formally conforms to our proposed definition of first-order preference (each element of $X_{oh}$ is multiplied by the corresponding element of $M_s$), 
in other word, element of $M_s$ reveal how important the corresponding elements in $X_{oh}$ is. 
We think it can reveal the preference of the scoring function at each position. 
For instance, follow the encoding rules above, train spatial domain part of AttCRISPR with the WT-SpCas9 dataset. 
Then take the average of all spatial attention matrix, and the element in the first row and third column are closer to 1, which means when calculate the final score, 
G typically may have a important contribution at first position within sgRNA sequence. 
In fact, this corresponds to some early studies concerning the Human (hU6) promoter, which is believed to require G as the first nucleotide of its transcript \citep{cong2013multiplex,jinek2012a,mali2013rnaguided}.

@@

    \centerline{\includegraphics[scale=0.15]{CNNv2.png}}

@@

y=\sum_i^lW_i\tilde{A}_i\cdot {X_e}_i\label{eq:12}\vspace*{-10pt}
\end{equation}\\

@@

Attention(Q,K,V)=align(Q,K)V\label{eq:13}\vspace*{-10pt}
\end{equation}\\

@@

\end{matrix}\right.\label{eq:14}\vspace*{-10pt}
\end{equation}\\

@@

\end{matrix}\right.\label{eq:18}\vspace*{-10pt}
\end{equation}\\

@@

\tilde{A}=BV\label{eq:19}\vspace*{-10pt}
\end{equation}\\

@@

Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.

Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
above method  Text Text Text Text Text Text Text Text Text Text
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text\break Text.

Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text\vspace*{-10pt}

>>>>

\text{\it y}=Score(\it X)\label{eq:01}

@@

dy = \sum^{d}_ia_idx_i\label{eq:02}

@@

a_i=\frac{\partial}{\partial x_i} Score(X)\label{eq:03}

@@

y=\sum_i^lW_iA_i\cdot X_i\label{eq:04}

@@

\tilde{A}=BA\label{eq:05}

@@

\tilde{A}_i=\sum_j^lB_{ij}A_j\label{eq:06}

@@

d\tilde{A}_i=\sum_j^lB_{ij}EdA_j\label{eq:07}

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[width=86mm]{category.png}}
    \caption{Two categories of the deep learning models used in sgRNA related task. 
    (a) Model work in spatial domain. 
    In spatial domain, the base sequence is encoded into a binary matrix (or a binary image). 
    Since convolution has great advantages in extracting spatial features, the convolutional neural network is an excellent tool in spatial domain. 
    (b) Model work in temporal domain. 
    In temporal domain, the base sequence (represented by the binary matrix) is embedded into a sequance of high-dimensional vector, 
    in which the recurrent neural network perform better.
    In addition, we note that the last layers of neural network are usually full connection structure (not necessarily), 
    which greatly increases the difficulty of understanding the decisions of model.
    }\label{fig:01}
\end{figure}

 where $E$ is the identity matrix, and we define $B$ as the second-order preference matrix, 
 it can explain how a particular pattern containing two nucleotides affects the base sequence, 
 for example, how does the degree of preference for the G adjacent to PAM be affected by a T at 2 bases upstream?
 The previous work (even based on conventional machine learning algorithms with high interpretability) 
 was limited to dimers which can only work on the adjacent base pairs \citep{Liu2019} 
 or have to perform complex feature engineering \citep{MuhammadRafid2020}.
 In light of the above, we believe it is essential to develope a model which can not only match deep learning based model in performance, 
 but also be comparable to conventional machine learning algorithms in interpretability.

@@

Attention mechanism has demonstrated its power in Natural Language Processing, Statistical Learning, Speech and Computer Vision \citep{chaudhari2019an}. 
 It makes model tends to focus selectively on parts of the input which is help in performing the task effectively. 
 Previous observation have shown that Cas9 preferentially binds sgRNAs containing purines but not pyrimidines \citep{wang2014genetic} and multiple thymine in the spacer impairing sgRNA activity \citep{wu2014genome-wide}, 
 that is to say some specific nucleotide and base position need more attention compared to others. 
 The above is the premise of introducing attention mechanism. Strictly speaking, we are not the first to bring attention mechanisms into this field. 
 The most similar approach to ours is the work based on transformer by \citeauthor{Liu2019}. 
 They use transformer, a components based on attention mechanism, instead of RNN to improve the ability of temporal feature extraction, 
 hence, enhance the performance of their model \citep{Liu2019,vaswani2017attention}. 
 In our work,the interpretability benefit from attention mechanism is more focused.
 In this paper, our main contributions are as follows:\vspace*{1pt}
\begin{itemize}
    \item Present a novel deep-learning-based model, which can extract potential feature representation of sgRNA sequence in both spatial and temporal domain parallelly. 
    Finally, the ensemble learning method is used to combine the two to achieve better performance than other models.
    It doesn't belong to any of the above categories of existing approaches.
    \item Introduce attention mechanism into our model. 
    As a result, It does not need post hoc explanations techniques based on input perturbation to explain itself. 
    It is intrinsic interpretable in both temporal and spatial domains.
    In the spatial domain it's at the nucleotide level, in our word, first-order preference, 
    while at the overall level in the temporal domain, in our word, second-order preference. 
    Thus, it is transformed from a black box to an intrinsically interpretable model with the performance of deep learning based model. 
    \item Throught abation analysis and testing a series of possible network structure, 
    we find there are multiple components and strategy can improve the performance of our model and constructed AttCRISPR, 
    which could outperform the current state-of-the-art tool on DeepHF dataset.\vspace*{1pt}
\end{itemize}

@@

To facilitate the training process, we can mapping each one-hot vector into a dense real-valued high-dimensional space, which is called embedding. 
Concretely speaking, at the matrix level, the formula is as follows
\begin{equation}
X_e=X_{oh}E_m\label{eq:08}
\end{equation}
where $X_e$ named embedding matrix, $E_m\in\mathbb{R}^{4\times m}$ is a trainable transformational matrix, $m$ refers to the dimension of embedding space. 
We believe it is also meaningful to regard nucleotides in the sgRNA sequence as word, and the sgRNA sequence itself as a sentence, 
Guided by this belief $E_m$ is the word embedding matrix and $X_e$ is the sentence embedding in natural language processing (NLP).
Therefore, $X_e$ is used as an input of recurrent neural network (or its variant), which performs well in the NLP field.
In the area of NLP, there have been many methods based on pre-training and unsupervised learning been developed to determine the $E_m$. 
However, we believe that sgRNA sequence is not a natural language in the true sense, the method above does not apply to it. 
So $E_m$ will be trained along with the model as a whole.

@@

\begin{figure}[!tpb]%figure2
    \centerline{\includegraphics[width=86mm]{CNNv2.png}}
    \caption{The architecture of spatial domain method in AttCRISPR. 
    The input of the method is encoded sgRNA sequence, a $21\times 4$ one-hot matrix. 
    Then refine it through a spatial attention module, which could tell us the importance of a specific matrix element (or just say, pixel).
    Finally, a simple convolutional neural network is applied.}\label{fig:02}
\end{figure}
As Figure~2\vphantom{\ref{fig:02}} show, method in spatial domain relies on the convolutional neural network. 
As previously mentioned, sgRNA sequence has been encoded into a $21\times 4$ one-hot matrix $X_{oh}$, and we regard $X_{oh}$ as a binary image. 
Then, convolution kernels with different size are used to extract potential spatial features just like what other do in computer vision. 
According to the foregoing, the Spatial Attention Module proposed by \citeauthor{woo2018cbam:} can be applied in our method in spatial domain, 
which was used to improve the performance of CNN in vision tasks.

\begin{figure}[!tpb]%figure3
    \centerline{\includegraphics[width=86mm]{spatialmodule.png}}
    \caption{Details of spatial attention module. A convolution layer is used to generate multi-channel map from $X_{oh}$.
    Then concatenated the output of both max-pooling and average-pooling method and forward it to the last convolution layer.
    A sigmoid is used to map the final result to a range of zero to one.
    }\label{fig:03}
\end{figure}

As Figure~3\vphantom{\ref{fig:03}} show, Given an one-hot map $X_{oh}$, spatial attention module generating spatial attention matrix 
$A_s\in\mathbb{R}^{l\times 4}$ with the same as $X_{oh}$ in shape. 
Each element of $A_s$ is constrained to a range of zero to one, implemented by a sigmoid function, which reflects the importance of the corresponding elements of $X_{oh}$.
The overall spatial attention process can be summarized as:
\begin{equation}
\left\{\begin{matrix}
X_{mc} = f^{3\times4}(X_{oh})
\\A_s = \sigma(f^{3\times2}([AvgPool(X_{mc});MaxPool(X_{mc})]))
\\X_{rf} = A_s\otimes X_{oh}
\end{matrix}\right.\label{eq:11}
\end{equation}
where $f^{p\times q}$ represents a convolution operation with the filter size of $p\times q$, 
$p,q\in\mathbb{Z}^{+}$, $X_{mc}$ is a multi-channel map generated by $X_{oh}$, $\sigma$ denotes the sigmoid function, 
AvgPool denotes the average-pooling operation, MaxPool denotes the max-pooling operation, $\otimes$ denotes element-wise multiplication.

Spatial attention matrix $A_s$ formally conforms to our proposed definition of first-order preference (each element of $X_{oh}$ is multiplied by the corresponding element of $A_s$), 
in other word, element of $A_s$ reveal how important the corresponding elements in $X_{oh}$ is. 
We think it can reveal the preference of the scoring function at each position. 
For instance, follow the encoding rules above, train spatial domain part of AttCRISPR with the WT-SpCas9 dataset. 
Then take the average of all spatial attention matrix, and the element in the first row and third column are closer to 1, which means when calculate the final score, 
G typically may have a important contribution at first position within sgRNA sequence. 
In fact, this corresponds to some early studies concerning the Human (hU6) promoter, which is believed to require G as the first nucleotide of its transcript \citep{cong2013multiplex,jinek2012a,mali2013rnaguided}.

@@

    \centerline{\includegraphics[width=86mm]{temporalmodule.png}}

@@

y=\sum_i^lW_i\tilde{A}_i\cdot {X_e}_i\label{eq:12}

@@

Attention(Q,K,V)=align(Q,K)V\label{eq:13}

@@

\end{matrix}\right.\label{eq:14}

@@

\end{matrix}\right.\label{eq:18}

@@

\tilde{A}=BV\label{eq:19}

@@
